---
title: "Data Exploration"
author: 'Kathy Lin'
date: "March 27, 2018"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE, 
                      echo=FALSE,
                      cache = TRUE)
```

First we will import the necessary packages for exploratory analysis.
```{r}
library(readr)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggthemes)
library(dplyr)
library(GGally)
library(vcd)
library(extracat)
library(DAAG)
library(forcats)
library(tibble)
library(lubridate)
library(skimr)
library(plotly)
library(parcoords)
```

Given these packages, we first do some missing value analysis and then we begin by trying to produce univariate graphs of the variables in our dataset over time. Note we've converted the song duration from ms to minutes 
```{r}
spotifydf <- read.csv("billboard-spotify.csv", stringsAsFactors=FALSE)
spotifydf <- as.tibble(spotifydf)
skimr::skim(spotifydf)
visna(spotifydf, sort='b')
```

There seems to be 337 songs that have 15 continuous features such as acousticness, danceability, duration, energy, explicitness, etc. missing at the same time. These songs will be removed for univariate analysis. These 337 NAs were well spreadout throughout the years.
```{r}
spotifyNA <- spotifydf%>%
            filter(is.na(acousticness))
```

To begin the univariate analysis, we converted duration_ms to duration in minutes for ease of comprehension.

Observations: 

* There is no clear overall trend in **duration** over the years
* There is a clear decline in **acousticness** of songs over the years of our dataset
* There is a surprising dip in **danceability** in 2010 -- Wonder what spiked in 2010 that caused this tradeoff
* There seems to be a general upward trend seen in **energy** throughout the years but the data also shows around 3 plateaus every 20 years. 
* There is a general decline in **instrumentalness** of songs over the years
* There is no clear trend in **liveness** of songs over the years. Liveness does peak at 2010 though
* There is a general upward increasing trend in **loudness** of songs over the years. There seems to be a peak around 1983 which may require further investigation why that is. Note negative scale for loudness makes sense
* From 1965-1990, there was a pretty constant level of speechiness in songs. **Speechiness** peaks around 2000-2005 -- investigate why that is
* Tempo remains fairly consistent around 115 during the years 1985-2005. And there is a dip in tempo in 2010.
* Valence generally declines over the years in our dataset. America was great...back when Trump wasn't president. LOL



```{r}
# Univariate Plots
spotifydf<-spotifydf%>%
    filter(!is.na(duration_ms))%>%
    mutate(duration=duration_ms/1000/60)


### Univariate Plots
spotifydf <- spotifydf%>%
                group_by(year) 
#Duration    
spotifydf2<- spotifydf%>%
                 summarize(average_duration = mean(duration))

g1 <- ggplot(spotifydf2, aes(year, average_duration)) + geom_line() +
    ggtitle("Average Duration over Time")+
    ylab("Average Duration (in Minutes)")+
    scale_y_continuous(breaks = seq(2.5, 5, .5))+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Acousticness
spotifydf3<- spotifydf%>%
                 summarize(avg_acousticness = mean(acousticness))

g2 <- ggplot(spotifydf3, aes(year, avg_acousticness)) + geom_line() +
    ggtitle("Average Acousticness over Time")+
    ylab("Average Acousticness")+
    scale_x_continuous(breaks = seq(1960,2020,5))+
    scale_y_continuous(breaks = seq(0.0, 0.5, 0.05))
    
#Danceability
spotifydf4<- spotifydf%>%
                 summarize(avg_dance = mean(danceability))

g3 <- ggplot(spotifydf4, aes(year, avg_dance)) + geom_line() +
    ggtitle("Average Danceability over Time")+
    ylab("Average Danceability ")+
    scale_x_continuous(breaks = seq(1960,2020,5))+
    scale_y_continuous(breaks = seq(0.40, 0.8, 0.05))

#Energy
spotifydf5<- spotifydf%>%
                 summarize(avg_energy = mean(energy))

g4 <- ggplot(spotifydf5, aes(year, avg_energy)) + geom_line() +
    ggtitle("Average Energy over Time")+
    ylab("Average Energy ")+
    scale_x_continuous(breaks = seq(1960,2020,5))+
    scale_y_continuous(breaks = seq(0.40, 0.8, 0.05))

#Instrumentalness
spotifydf6<- spotifydf%>%
                 summarize(avg_instrument = mean(instrumentalness))

g5 <- ggplot(spotifydf6, aes(year, avg_instrument)) + geom_line() +
    ggtitle("Average Instrumentalness over Time")+
    ylab("Average Instrumentalness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Liveness
spotifydf7<- spotifydf%>%
                 summarize(avg_liveness = mean(liveness))

g6 <- ggplot(spotifydf7, aes(year, avg_liveness)) + geom_line() +
    ggtitle("Average Liveness over Time")+
    ylab("Average Liveness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))+
    scale_y_continuous(breaks = seq(0.0, 0.3, 0.02))

#Loudness
spotifydf8<- spotifydf%>%
                 summarize(avg_loudness = mean(loudness))

g7 <- ggplot(spotifydf8, aes(year, avg_loudness)) + geom_line() +
    ggtitle("Average Loudness over Time")+
    ylab("Average Loudness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Speechiness
spotifydf9<- spotifydf%>%
                 summarize(avg_speech = mean(speechiness))

g8 <- ggplot(spotifydf9, aes(year, avg_speech)) + geom_line() +
    ggtitle("Average Speechiness over Time")+
    ylab("Average Speechiness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))+ 
    scale_y_continuous(breaks = seq(0.0,0.15,0.03))

#Tempo
spotifydf10<- spotifydf%>%
                 summarize(avg_tempo = mean(tempo))

g9 <- ggplot(spotifydf10, aes(year, avg_tempo)) + geom_line() +
    ggtitle("Average Tempo over Time")+
    ylab("Average Tempo ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Valence
spotifydf11<- spotifydf%>%
                 summarize(avg_valence = mean(valence))

g10 <- ggplot(spotifydf11, aes(year, avg_valence)) + geom_line() +
    ggtitle("Average Valence over Time")+
    ylab("Average Valence")+
    scale_x_continuous(breaks = seq(1960,2020,5))


grid.arrange(g1,g2)
grid.arrange(g3,g4) 
grid.arrange(g5,g6)
grid.arrange(g7,g8)
grid.arrange(g9,g10)
```
Since plotting the averages gives us limited insight to the behavior of these variables over time, we now plot boxplots of their distributions over all years

```{r}
### Univariate Plots
spotifydf <- spotifydf%>%
                group_by(year) 
#Duration    
ggplot(spotifydf) + geom_boxplot(aes(year, duration, group=year))+
    ggtitle("Duration over Time")+
    ylab("Duration (in Minutes)")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Acousticness
ggplot(spotifydf) + geom_boxplot(aes(year, acousticness, group=year)) +
    ggtitle("Acousticness over Time")+
    ylab("Acousticness")+
    scale_x_continuous(breaks = seq(1960,2020,5))
    
#Danceability
ggplot(spotifydf) + geom_boxplot(aes(year, danceability, group=year)) +
    ggtitle("Danceability over Time")+
    ylab("Danceability ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Energy
ggplot(spotifydf) + geom_boxplot(aes(year, energy, group=year)) +
    ggtitle("Energy over Time")+
    ylab("Energy ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Instrumentalness
ggplot(spotifydf) + geom_boxplot(aes(year, instrumentalness, group=year)) +
    ggtitle("Instrumentalness over Time")+
    ylab("Instrumentalness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Liveness
ggplot(spotifydf) + geom_boxplot(aes(year, liveness, group=year)) +
    ggtitle("Liveness over Time")+
    ylab("Liveness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Loudness
ggplot(spotifydf) + geom_boxplot(aes(year, loudness, group=year)) +
    ggtitle("Loudness over Time")+
    ylab("Loudness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Speechiness
ggplot(spotifydf) + geom_boxplot(aes(year, speechiness, group=year)) +
    ggtitle("Speechiness over Time")+
    ylab("Speechiness ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Tempo
ggplot(spotifydf) + geom_boxplot(aes(year, tempo, group=year)) +
    ggtitle("Tempo over Time")+
    ylab("Tempo ")+
    scale_x_continuous(breaks = seq(1960,2020,5))

#Valence
ggplot(spotifydf) + geom_boxplot(aes(year, valence, group=year)) +
    ggtitle("Valence over Time")+
    ylab("Valence")+
    scale_x_continuous(breaks = seq(1960,2020,5))


```

## Parallel Coordinate Plots

Now that we've plotted the continuous variables univariately, we'll now plot them on the same time series graph to observe patterns or correlations in their behavior. For convenience, we've rescaled the variables such that the ggparcoord graph makes sense. 

```{r}
#Define subsetted spotify df with keepcols
keepcols <- c('year', 'acousticness', 'danceability', 'duration', 'energy', 'instrumentalness', 'liveness','loudness','speechiness', 'tempo', 'valence')

spotifydf_s <- spotifydf%>%
                select(keepcols) %>%
                group_by(year)%>%
                summarize(mean_acousticness = mean(acousticness), 
                        mean_danceability = mean(danceability), 
                        mean_duration = mean(duration), 
                        mean_energy = mean(energy), 
                        mean_instrumentalness = mean(instrumentalness), 
                        mean_liveness = mean(liveness), 
                        mean_loudness = mean(loudness), 
                        mean_speechiness = mean(speechiness), 
                        mean_tempo = mean(tempo), 
                        mean_valence = mean(valence)
                )
# %>%
#                 gather(key='variable', value = 'Freq', -year)
spotifydf_s$year<- factor(spotifydf_s$year, levels = unique(spotifydf_s$year))

ggparcoord(spotifydf_s, columns = 2:11, alphaLines = 0.7, groupColumn ='year',scale = 'uniminmax')+xlab("")+ylab("") + theme(axis.text.x = element_text(angle=90)) 

# custom function to transpose while preserving names
transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  return(t_df)
}

# using the function
dft<-transpose_df(spotifydf_s)
colnames(dft) <- dft[1, ]
dft <- dft[-1,]
dft<- dft%>%
    rownames_to_column('Variable')

ggparcoord(dft, columns = 2:52, alphaLines = 0.7, groupColumn ='Variable',scale = 'uniminmax')+xlab("")+ylab("") + theme(axis.text.x = element_text(angle=90)) 
```


We will now attempt interactive parallel coordinate plots. 
```{r}
#devtools::install_github("timelyportfolio/parcoords")


spotifydf_s$year <- year(as.Date(as.character(spotifydf_s$year), "%Y"))

rownames(spotifydf_s)<- spotifydf_s$year
p <- spotifydf_s%>% arrange(year) %>%
    parcoords(
        rownames = F,
        brushMode = "1D-axes",
        reorderable = T,
        queue = T, 
        alpha=.8, 
        color = list(colorBy = "year", colorScale = htmlwidgets::JS("d3.scale.category10()")), 
        width = 1100, 
        height = 500
    )
p

```